Using Two-Level Type Theory for Staged Compilation \thanks{The author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}

### 2021-04-22 types2021/abstract.tex
update abstract

The idea of staging is the following: given a closed $A : \U^0$ with a closed $t
: A$ in 2LTT, there are unique $A'$ and $t'$ in the object theory, which become
definitionally equal to $A$ and $t$ respectively after being embedded in
2LTT. In short, every meta-level construction can be computed away. Annekov et
al. \cite{twolevel} showed this without uniqueness of $A'$ and $t'$. We can get
uniqueness as well, using the normalization of 2LTT: by induction on the
(unique) normal forms of $A$ and $t$, we can show that they cannot contain
meta-level subterms. This shows that normalization is a sound staging algorithm,
but in practice we do not want to compute full normal forms; we want to compute
\emph{meta-level redexes only}. This can be done with a variation of standard
normalization-by-evaluation \cite{abel2013normalization,
  Wieczorek:2018:CFN:3176245.3167091} which also keeps track of stages.

@@

\textbf{Control over lambda lifting and closure creation}. We assume now a
dependent type theory on both levels, but with a \emph{first-order function
type} on the object level. This is defined by splitting $\U^0$ to a universe
$\V^0$ which is closed under inductive types but not functions, and a universe
$\C^0$ which has $V^0$ as a sub-universe, and is closed under functions with
domains in $\V^0$ and codomains in $\C^0$. The object theory supports
compilation which requires only lambda lifting, but no closures. On its own, the
object theory is fairly restricted, but together with staging we have a
remarkably expressive system. Then, we can close $\V^0$ under a separate type
former of closure-based functions, thereby formally distinguishing
lambda-liftable functions from closure-based functions. This enables typed
analysis of various optimization and fusion techniques. E.g.\ we get guaranteed
closure-freedom in code output if a certain fusion technique can be formalized
with only first-order function types. In particular, this may obviate the need
for arity analysis \cite{DBLP:journals/cl/Breitner18} in fold-based fusion.

>>>>

The idea of staging is the following: given a closed $A : \U^0$ with a closed $t
: A$ in 2LTT, there are unique $A'$ and $t'$ in the object theory, which become
definitionally equal to $A$ and $t$ respectively after being embedded in
2LTT. In short, every meta-level construction can be computed away, and only
object-level constructions remain in the result. Annekov et al. \cite{twolevel}
showed this without uniqueness of $A'$ and $t'$ (as a conservativity theorem for
2LTT). We can get uniqueness as well, using the normalization of 2LTT: by
induction on the (unique) normal forms of $A$ and $t$, we can show that they
cannot contain meta-level subterms. This shows that normalization is a sound
staging algorithm, but in practice we do not want to compute full normal forms;
we want to compute \emph{meta-level redexes only}. This can be done with a
variation of standard normalization-by-evaluation \cite{abel2013normalization,
  Wieczorek:2018:CFN:3176245.3167091} which also keeps track of stages.

@@

\textbf{Control over lambda lifting and closure creation}. We assume now a
dependent type theory on both levels, but with a \emph{first-order function
type} on the object level. This is defined by splitting $\U^0$ to a universe
$\V^0$ which is closed under inductive types but not functions, and a universe
$\C^0$ which has $V^0$ as a sub-universe, and is closed under functions with
domains in $\V^0$ and codomains in $\C^0$. This object theory supports
compilation which requires only lambda lifting, but no closures. On its own, the
object theory is fairly restricted, but together with staging we have a
remarkably expressive system. Then, we can close $\V^0$ under a separate type
former of closure-based functions, thereby formally distinguishing
lambda-liftable functions from closure-based functions. This enables typed
analysis of various optimization and fusion techniques. E.g.\ we get guaranteed
closure-freedom in code output if a certain fusion technique can be formalized
with only first-order function types. In particular, this may obviate the need
for arity analysis \cite{DBLP:journals/cl/Breitner18} in fold-based fusion.
